---
title: "Understanding Variability via Simulations"
author: "Brian Baquiran"
date: "November 17, 2014"
output: html_document
---
# Simulation Example #1

Standard normals have variance 1; the _means_ of $n$ samples of a standard normal have a standard error $\frac{1}{\sqrt{n}}$

```{r}
nosim <- 1000
n <- 10
```

Let's simulate `r nosim` draws of `r n` samples from a standard normal population. 
For each simulation we get the mean, and then the standard deviation of all of the `r nosim` means.

```{r}
sd(apply(matrix(rnorm(nosim * n), nosim), 1, mean))
```

We can compare this value to the $\frac{1}{\sqrt{n}}$ value we expect for the standard error.

```{r}
1/sqrt(n)
``` 

Note that it's close to the random value above.

## Quiz 2

### Question 1
What is the variance of the distribution of the average an IID draw of n observations from a population with mean $\mu$ and variance $\sigma^2$.

The variance of the sample mean is $\frac{\sigma^2}{n}$.

### Question 2
Suppose that diastolic blood pressures (DBPs) for men aged 35-44 are normally distributed with a mean of 80 (mm Hg) and a standard deviation of 10. About what is the probability that a random 35-44 year old has a DBP less than 70?

Note that 70 is one standard deviation lower than the mean. Recall that 68% of the observations will be within one standard deviation, and half of that (34%) will be below the mean. Since in the normal distribution the mean and median are the same, we can just subtract 34% from 50% and get 16%. So the probability that the random subject will have a DBP one standard deviation lower than the mean, is 16%.

To solve this in R, we use _pnorm_
```{r}
mean <- 80
sd <- 10
DBP <- 70
pnorm(DBP,mean,sd)
```

### Question 3
Brain volume for adult women is normally distributed with a mean of about 1,100 cc for women with a standard deviation of 75 cc. About what brain volume represents the 95th percentile?

To solve this in R, we use _qnorm_
```{r}
mean <- 1100
sd <- 75
p <- 0.95
qnorm(p,mean,sd)
```

### Question 4
Refer to the previous question. Brain volume for adult women is about 1,100 cc for women with a standard deviation of 75 cc. Consider the sample mean of 100 random adult women from this population. Around what is the 95th percentile of the distribution of that sample mean?

The question refers to the distribution of the sample mean. The sample mean has a standard error of s/sqrt(n). To get the 95th percentile of sample means, we just use qnorm.

```{r}
n <- 100
standard_dev <- 75
standard_error <- standard_dev/sqrt(100)
qnorm(.95,standard_error,mean=1100)
```

### Question 5
You flip a fair coin 5 times, about what's the probability of getting 4 or 5 heads?

This is a Binomial distribution. We can compute the probabilities for 4 heads and 5 heads separately, then add.
```{r}
choose(5,4) * .5 ^ 5 + choose(5,5) * .5 ^ 5
```

Alternately, we could use the pbinom to compute for the possibility.
```{r}
pbinom(3,size=5,prob=.5,lower.tail = FALSE)
```

### Question 6
The respiratory disturbance index (RDI), a measure of sleep disturbance, for a specific population has a mean of 15 (sleep events per hour) and a standard deviation of 10. They are not normally distributed. Give your best estimate of the probability that a sample mean RDI of 100 people is between 14 and 16 events per hour?

Again, we are interested in the distribution of sample means. Thus we need the sample error. 
```{r}
RDI_mean <- 15
RDI_sd <- 10
RDI_n <- 100
RDI_sample_err <- 10/sqrt(100)
```
Our sample error is `r RDI_sample_err`.

Even if the individual observations are not normally distributed, CLT says that the means of samples will be normally distributed. The range we're interested in is exactly one standard error away from the mean, so 68% of the sample means will be within this range.

### Question 7
Consider a standard uniform density. The mean for this density is .5 and the variance is $\frac{1}{12}$. You sample 1,000 observations from this distribution and take the sample mean, what value would you expect it to be near?

It will be near the mean, 0.5.

### Question 8
The number of people showing up at a bus stop is assumed to be Poisson with a mean of 5 people per hour. You watch the bus stop for 3 hours. About what's the probability of viewing 10 or fewer people?

```{r}
ppois(10, lambda=5*3)
```

## Quiz 3

Student's t-test, confidence intervals, hypothesis testing. 

### Question 1
In a population of interest, a sample of 9 men yielded a 
sample average brain volume of 1,100cc and a standard 
deviation of 30cc. What is a 95% Student's T confidence 
interval for the mean brain volume in this new population?

```{r}
df <- 9-1
mu <- 1100
s <- 30

# Use dt() to get the density
1100 + c(-1,1) * qt(.975,8) * 30/sqrt(9)
```

### Question 2
A diet pill is given to 9 subjects over six weeks. The average difference in weight (follow up - baseline) is -2 pounds. 
What would the standard deviation of the difference in 
weight have to be for the upper endpoint of the 95% T 
confidence interval to touch 0?

* 2.10
* 2.60
* 0.30
* 1.50


The T interval is $\bar{x}\pm t_{n-1} \frac{s}{\sqrt{n}}$. 
```{r}
n <- 9
df <- n - 1
xbar <- -2
s <- -xbar * sqrt(n) / qt(.975,df=8)
print(s)
```

### Question 3
In an effort to improve running performance, 5 runners were either given a protein supplement or placebo. Then, after a suitable washout period, they were given the opposite treatment. Their mile times were recorded under both the treatment and placebo, yielding 10 measurements with 2 per subject. The researchers intend to use a T test and interval to investigate the treatment. Should they use a paired or independent group T test and interval?

*Paired*

### Question 4
In a study of emergency room waiting times, investigators consider a new and the standard triage systems. To test the systems, administrators selected 20 nights and randomly assigned the new triage system to be used on 10 nights and the standard system on the remaining 10 nights. They calculated the nightly median waiting time (MWT) to see a physician. The average MWT for the new system was 3 hours with a variance of 0.60 while the average MWT for the old system was 5 hours with a variance of 0.68. Consider the 95% confidence interval estimate for the differences of the mean MWT associated with the new system. Assume a constant variance. What is the interval? Subtract in this order (New System - Old System).

* __[-2.75, -1.25]__
* [1.29, 2.70]
* [1.25, 2.75]
* [-2,70, -1.29]

We used the formula for t confidence interval for independent groups.
For a (1-$\alpha$)x 100% confidence interval for $\mu_{y}-\mu_{z}$ this is:
$$\bar{Y}-\bar{X}\pm t_{n_{z}+n_{y}-2,1-\alpha/2} S_{p}\left(\frac{1}{n_{z}}+\frac{1}{n_{y}}\right)^{1/2}$$

where $S_p^2$ is the pooled variance estimator, the average of the variance of both groups weighted by the number of observations in each group:

$$S_p^2 = \{(n_z - 1)S_z^2+(n_y-1)S_y^2\}/(n_z+n_y-2)$$

$S_z^2$ and $S_y^2$ are the 0.60 and 0.68 variances provided.
```{r}
MWT.new <- 3
Var.new <- 0.60
MWT.old <- 5
Var.old <- 0.68
n.new <- 10
n.old <- 10
# Degrees of freedom
df <- n.new+n.old-2
alpha <- 5

# pooled standard deviation
S.p <- sqrt(((n.new - 1)*Var.new + (n.old-1)*Var.old)/df)

# Confidence interval for independent groups
(MWT.new - MWT.old) + c(-1,1) * qt(.975,df) * S.p * sqrt(1/n.new + 1/n.old)

```

### Question 5
Suppose that you create a 95% T confidence interval. You then create a 90% interval using the same data. What can be said about the 90% interval with respect to the 95% interval?

* The interval will be the same width, but shifted.
* __The interval will be narrower.__
* The interval will be wider
* It is impossible to tell.

### Question 6
To further test the hospital triage system, administrators selected 200 nights and randomly assigned a new triage system to be used on 100 nights and a standard system on the remaining 100 nights. They calculated the nightly median waiting time (MWT) to see a physician. The average MWT for the new system was 4 hours with a standard deviation of 0.5 hours while the average MWT for the old system was 6 hours with a standard deviation of 2 hours. Consider the hypothesis of a decrease in the mean MWT associated with the new treatment. What does the 95% independent group confidence interval with unequal variances suggest vis a vis this hypothesis? (Because there's so many observations per group, just use the Z quantile instead of the T.)

* __When subtracting (old - new) the interval is entirely above zero. The new system appears to be effective.__
* When subtracting (old - new) the interval contains 0. There is not evidence suggesting that the new system is effective.
* When subtracting (old - new) the interval is entirely above zero. The new system does not appear to be effective.
* When subtracting (old - new) the interval contains 0. The new system appears to be effective.
```{r}
MWT.new <- 4
s.new <- 0.5
MWT.old <- 6
s.old <- 2
n.new <- 100
n.old <- 100
# Degrees of freedom
df <- n.new+n.old-2
alpha <- 5

# pooled standard deviation
S.p <- sqrt(((n.new - 1)*s.new^2 + (n.old-1)*s.old^2)/df)

# Confidence interval for independent groups
(MWT.old-MWT.new) + c(-1,1) * qnorm(.975) * S.p * sqrt(1/n.new + 1/n.old)
```
### Question 7
Suppose that 18 obese subjects were randomized, 9 each, to a new diet pill and a placebo. Subjects’ body mass indices (BMIs) were measured at a baseline and again after having received the treatment or placebo for four weeks. The average difference from follow-up to the baseline (followup - baseline) was −3 kg/m2 for the treated group and 1 kg/m2 for the placebo group. The corresponding standard deviations of the differences was 1.5 kg/m2 for the treatment group and 1.8 kg/m2 for the placebo group. Does the change in BMI over the four week period appear to differ between the treated and placebo groups? Assuming normality of the underlying data and a common population variance, calculate the relevant *90%* t confidence interval. Subtract in the order of (Treated - Placebo) with the smaller (more negative) number first.

* [2.636, 5.364]
* [2.469, 5.531]
* [-5.531, -2.469]
* __[-5.364, -2.636]__

```{r}
Xbar <- (-3)
s.x <- 1.5
Ybar <- 1
s.y <- 1.8
n.y <- n.x <- 9
alpha <- .10
df <- n.x + n.y - 2

# pooled standard deviation
S.p <- sqrt(((n.x - 1)*s.x^2 + (n.y-1)*s.y^2)/df)
print(S.p)  

# Confidence interval for independent groups
(Xbar-Ybar) + c(-1,1) * qt(1-alpha/2,df) * S.p * sqrt(1/n.x + 1/n.y)
```
