---
title: "Sample Means of an Exponential Distribution"
author: "Brian Baquiran"
date: "November 22, 2014"
output: html_document
---
## The Exponential Distribution

The exponential distribution represents the distribution of 
waiting times in a Poisson process with rate of change $\lambda$. 

The probability density function for the exponential distribution is 
$$f(x;\lambda)=\begin{cases}
\lambda e^{-\lambda x} & x \ge 0 \\
0 & x<0
\end{cases}
$$

The distribution is supported on the interval $[0, \infty)$. 

The mean of an exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also also $\frac{1}{\lambda}$.

In R the exponential distribution is simulated with rexp(n, lambda) where lambda is the rate parameter.
```{r}
n <- 40
lambda <- 0.2

# our expected mean and standard deviation
1/lambda
```

We expect to observe a mean and standard deviation of `r 1/lambda` from our simulation. 

## Simulation
To get a more intuitive understanding of the exponential 
distribution, let us first see what it looks like. We ask R 
to give us 40,000 random values from the distribution, and 
plot them on a histogram.

```{r}
# 1,000 simulations
num.sims <- 1000
X <- rexp(n*num.sims,lambda)
hist(X,breaks=100, main = "Exponential Distribution")
```

### Mean and Standard Deviation
We compute the mean and standard deviation of the random values 
and compare them to our expected value and standard deviation 
of $\frac{1}{\lambda}$.
```{r}
print(Xbar<-mean(X))
```
The mean for the 40,000 generated random values is `r Xbar`.

```{r}
sd(X)
```

The standard deviation for the random values is `r sd(X)`.

Both the observed mean and standard deviation are close to 
$\frac{1}{\lambda}$ which we know to be `r 1/lambda`.

```{r}
X2 <- matrix(X,nrow=num.sims)
d_means <- apply(X2,1,mean)
hist(d_means,breaks=50, main ="Means of sample size 40")
```
