---
title: "Sample Means of an Exponential Distribution"
author: "Brian Baquiran"
date: "November 22, 2014"
output: html_document
---

The exponential distribution represents the distribution of 
waiting times in a Poisson process with rate of change $\lambda$. 
The probability density function for the exponential distribution is 
$$f(x;\lambda)=\begin{cases}
\lambda e^{-\lambda x} & x \ge 0 \\
0 & x<0
\end{cases}
$$

The distribution is supported on the interval $[0, \infty)$. 
The mean of an exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also also $\frac{1}{\lambda}$.

In R the exponential distribution is simulated with rexp(n, lambda) where lambda is the rate parameter.
```{r}
n <- 40
lambda <- 0.2

# our expected mean and standard deviation
mu <- 1/lambda
sigma <- 1/lambda
```

We expect to observe a mean $\mu=5$ and standard deviation of $\sigma=5$ from our simulation. 

### Simulation
To get a more intuitive understanding of the exponential 
distribution, let us first see what it looks like. We ask R 
to give us 40,000 random values from the distribution, and 
plot them on a histogram. The distribution mean of `r mu` is indicated 
by the vertical line.

```{r}
library("ggplot2")
# 1,000 simulations
num.sims <- 1000
X <- rexp(n*num.sims,lambda)
p<-qplot(X,geom=c("histogram"),binwidth=1, main="Exponential Distribution, lambda=0.2") + geom_vline(aes(xintercept=1/lambda))
print(p)
```

### Mean and Standard Deviation
We compute the mean and standard deviation of the random values 
and compare them to our expected value and standard deviation 
of $\frac{1}{\lambda}$.
```{r}
print(Xbar<-mean(X))
```
The mean for the 40,000 generated random values is `r Xbar`.

```{r}
sd(X)
```

The standard deviation for the random values is `r sd(X)`.

Both the observed mean and standard deviation are close to 
$\frac{1}{\lambda}$ which we know to be `r 1/lambda`.

### Sample means for n=40. 
We divide the 40,000 random values we got earlier into 1000 
samples of n=40, and get the mean of each sample. The
expected value is indicated by a vertical line.
```{r}
X2 <- matrix(X,nrow=num.sims)
X2_means <- apply(X2,1,mean)
#hist(X_means,breaks=50, main ="Means of sample size 40")
p<-qplot(X2_means,geom=c("histogram"),
        main="Means of n=40, lambda=0.2") + 
        geom_vline(aes(xintercept=1/lambda))
print(p)
```

The distribution of the sample means _looks_ normal and is 
centered around the distribution mean of `r mu`. The mean of 
the sample means is in fact `r mean(X2_means)`.

### Standard error for n=40. 

The standard error of means of size n is $\frac{\sigma}{\sqrt{n}}$. 
 
```{r}
std.err = sigma/sqrt(n)
```
We find that for samples of size n=40, standard error is `r std.err`.
We can compare this to the standard deviation of the sample
means.
```{r}
sd(X2_means)
```

So this confirms our understanding that the standard deviation
of the sample means is estimated by the standard error 
$\frac{\sigma}{\sqrt{n}}$.