---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Brian Baquiran"
date: "November 23, 2014"
output: pdf_document
---
### The ToothGrowth Dataset
The response is the length of odontoblasts (teeth) in each 
of 10 guinea pigs at each of three dose levels of Vitamin C 
(0.5, 1, and 2 mg) with each of two delivery methods (orange
juice or ascorbic acid).

The data consists of a data frame with 60 observations on 3 
variables.

Index|colname|class|description
-----|-------|-----|-------------
[,1]|len|numeric|Tooth length
[,2]|supp|factor|Supplement type (VC or OJ).
[,3]|dose|numeric|Dose in milligrams.


### Exploratory Analysis
A box plot of the resulting tooth growth for different dosages 
and delivery methods of vitamin C:
```{r, echo=FALSE}
library("datasets"); library("plyr"); library("ggplot2")
data("ToothGrowth")
qplot(dose, len, colour=supp, data=ToothGrowth, facets= dose ~ supp, geom="boxplot")
```

A summary of the means and standard deviation and error for 
the samples of the different treatments.
```{r}
ddply(.data = ToothGrowth, .variables = .(supp,dose), summarize, mean = mean(len), std.dev = sd(len), n=length(len), std.error = sd(len)/sqrt(length(len)))
```

From the summary data and the box plot, it would appear that 
for a given dose (0.5mg, 1.0mg, 2.0mg) the OJ delivery method
results in greater tooth growth. 

### Hypothesis
Our null hypothesis is that the mean tooth growth for the 
orange juice (OJ) group is equal to the mean tooth growth of
the ascorbic acid group (VC), for the same dosage of vitamin 
C. The alternative hypothesis is that they are not equal.

$$H_{0}:\mu_{vc}=\mu_{oj} \Leftrightarrow \mu_{vc}-\mu_{oj} = 0$$
$$H_{a}:\mu_{vc}\neq\mu_{oj}$$

Our test statistic will be that for two group intervals:

$$TS=\frac{\overline{X_{1}}-\overline{X_{2}}}{\sqrt{\frac{S_{1}^2}{n_{1}}+\frac{S_{2}^2}{n_{2}}}}$$

### Assumptions
We will assume $\alpha=0.05$, or a 95% confidence interval. 
We will use the Student t distribution due to the small 
sample size (n=10 for each treatment modality). 

The rejection region can be obtained using qt() for our desired confidence interval:
```{r}
c(qt(0.025,9),qt(0.975,9))
```
We can reject $H_{0}$ if the test statistic falls outside 
this interval.

### Comparing OJ and VC for the different doses
We can use the mean and standard deviations for each treatment
modality computed earlier to determine the test statistic for
comparing OJ and VC for a 0.5mg vitamin C dose.
```{r}
(13.23-7.98)/sqrt((4.460^2/10)+(2.747^2)/10)
```
This test statistic falls outside the range [-2.262,2.262] so 
we can reject the null hypothesis for this dose. 

We could, alternately, have used t.test:
```{r}
t.test(ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5,]$len,ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5,]$len)$statistic
```
This matches our computation.

Performing the same test statistic computation for 1.0mg:
```{r}
(22.70-16.77)/sqrt((3.911^2/10)+(2.515^2)/10)
```
```{r}
t.test(ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0,]$len,ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0,]$len)$statistic
```
Our manual computation agrees with the t.test computation. 
We can again reject the null hypothesis for 1.0mg.

Finally, for 2.0mg we perform the same computation of the 
test statistic.
```{r}
(26.06-26.14)/sqrt((2.655^2/10)+(4.798^2)/10)
```
```{r}
t.test(ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2.0,]$len,ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2.0,]$len)$statistic
```
This time the test statistic falls within the interval
[-2.262,2.262] and we cannot reject the null hypothesis for 
2.0mg.

### Conclusions
From the data and analysis we conclude that we reject the 
null hypothesis $H_{0}:\mu_{VC}=\mu_{OJ}$ and adopt the 
alternative hypothesis $H_{a}:\mu_{VC}\neq\mu_{OJ}$ but only 
for dosages 0.5mg and 1.0mg. For 2.0mg, the data does not 
support rejecting the null hypothesis.